Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-03-22 manuscript/ms.tex
More edits.

As a first application of {\bf Ensembler}, we have built models for all 90 human tyrosine kinase (TK) domains listed in UniProt.
{\color{red}[JDC: What query was used to yield this set?]}
{\color{red}[JDC: Is there a complete list of these somewhere?  Maybe reference supplementary data?]}
TKs (and protein kinases in general) play important roles in many cellular processes and are involved in a number of types of cancer.
{\color{red}[JDC: CITE]}
For example, mutations of Src are associated with colon, breast, and prostate cancer {\color{red}[CITE: Src cancer involvement]}, while a translocation between the TK Abl1 and the pseudokinase Bcr is closely associated with chronic myelogenous leukemia {\color{red}[CITE: Abl1 cancer involvement]}.
Protein kinase domains are thought to have multiple accessible metastable conformation states, with a single active conformation, and much effort is directed at developing kinase inhibitor drugs which bind to and stabilize inactive conformations [CITE: Lee and Craik Science 2009].
{\color{red}[JDC: Lee and Craik do not discuss kinases, I don't believe; you'll have to find an accurate reference on kinase conformations.]}
Kinases are thus a particularly interesting subject for study with MSM methods [CITE: recent kinase MSM papers], and this approach stands to benefit greatly from the ability to exploit the full body of available genomic and structural data within the kinase family, e.g.~by generating large numbers of starting configurations to be used in highly parallel MD simulation.

We selected all available structures of protein kinase domains (of any species) as templates, for a total of 4433 (\num{398970} target-template pairs).
The templates were derived from 3028 individual PDB entries and encompassed 23 different species, with 3634 template structures from human kinase constructs.

@@

@@

To provide a more complete evaluation of the models generated, we have analyzed two example TKs (\emph{Src} and \emph{Abl1}) in detail.
Due to their importance in cancer, as outlined above, these kinases have been the subject of numerous studies, encompassing many different methodologies.
In terms of structural data, a large number of crystal structures have been solved (with or without ligands such as nucleotide substrate or inhibitor drugs), showing the kinases in a number of different conformations.
These two kinases are thus also interesting targets for MSM studies, with one recent study focusing on modeling the states which constitute the activation pathway of Src~\cite{shukla-pande:nature-commun:2014:src-activation-msm}.
Fig.~\ref{figure:superposition} shows a superposition of a set of representative models of \emph{Src} and \emph{Abl1}.
Models were first stratified into three ranges, based on the structure of the sequence identity distribution (Fig.~\ref{figure:sequence-identity-distribution}), then subjected to $k$-medoids clustering to pick three representative models from each sequence identity range.
{\color{red}[JDC: Explain how $k$-medoids clustering was done either here or in figure caption.]}
Each model is colored and given a transparency based on the sequence identity between the target and template sequence.
The figure gives an idea of the variance present in the generated models.
High sequence identity models (in opaque blue) tend to be quite structurally similar, with some variation in loops or changes in domain orientation.

The Abl1 renderings indicate one high sequence identity model with a long unstructured region at one of the termini, which was unresolved in the original template structure.
While such models are not necessarily incorrect or undesirable, it is important to be aware of the effects they may have on production simulations performed under periodic boundary conditions, as long unstructured termini can be prone to interact with a protein's periodic image.
Lower sequence identity models (in transparent white or red) indicate much greater variation in all parts of the structure.
We believe the mix of high and low sequence identity models to be particularly useful for methods such as MSM building, which require thorough sampling of the conformational landscape.
The high sequence identity models could be considered to be the most likely to accurately represent true metastable states.
Conversely, the lower sequence identity models could be expected to help push a simulation into regions of conformation space which might take intractably long to reach if starting a single metastable conformation.

To evaluate the models of \emph{Src} and \emph{Abl1} in the context of the published structural biology literature, we have focused on two residue pair distances thought to be important for the regulation of protein kinase domains.
We use the residue numbering schemes for chicken Src (which is commonly used in the literature even in reference to human Src)[CITE: 2SRC, 1Y57] and human Abl1 isoform A[CITE: 2F4J, 2HYY, 2G1T] respectively; the exact numbering schemes are provided in Supporting Information S\ref{si:residue-numbering-schemes}.
Fig.~\ref{figure:src-ref-structures} shows two structures of \emph{Src} believed to represent inactive (PDB code: 2SRC) {\color{red}[CITE: 2SRC]} and active (PDB code: 1Y57) {\color{red}[CITE: 1Y57]} states.
One notable feature which distinguishes the two structures is the transfer of an electrostatic interaction of E310 from R409 (in the inactive state) to K295 (in the active state), brought about by a rotation of the $\alpha$C-helix.
These three residues are also well conserved [CITE Kannan Neuwald JMB 2005], and a number of experimental and simulation studies have suggested that this electrostatic switching process plays a role in a regulatory mechanism shared across the protein kinase family~\cite{shukla-pande:nature-commun:2014:src-activation-msm} [CITE Foda Shan Seeliger Src Nat Commun 2015; Ozkirimli Post Prot Sci 2008].
As such, we have projected the {\bf Ensembler} models for \emph{Src} and \emph{Abl1} onto a space consisting of the distances between these two residue pairs (Fig.~\ref{figure:pair-distances}).
The models show strong coverage of regions in which either of the electrostatic interactions is formed, as well as a wide range of regions inbetween.
We thus expect that such a set of models, if used as starting configurations for highly parallel MD simulation, could greatly aid in sampling of the activation process.

>>>>

As a first application of {\bf Ensembler}, we have built models for all 90 human tyrosine kinase (TK) domains listed in UniProt.
{\color{red}[JDC: Is there a complete list of these somewhere?  Maybe reference supplementary data?]}
TKs (and protein kinases in general) play important roles in many cellular processes and are involved in a number of types of cancer.
{\color{red}[JDC: CITE]}
For example, mutations of Src are associated with colon, breast, and prostate cancer {\color{red}[CITE: Src cancer involvement]}, while a translocation between the TK Abl1 and the pseudokinase Bcr is closely associated with chronic myelogenous leukemia {\color{red}[CITE: Abl1 cancer involvement]}.
Protein kinase domains are thought to have multiple accessible metastable conformation states, with a single active conformation, and much effort is directed at developing kinase inhibitor drugs which bind to and stabilize inactive conformations [CITE: Lee and Craik Science 2009].
{\color{red}[JDC: Lee and Craik do not discuss kinases, I don't believe; you'll have to find an accurate reference on kinase conformations.]}
Kinases are thus a particularly interesting subject for study with MSM methods [CITE: recent kinase MSM papers], and this approach stands to benefit greatly from the ability to exploit the full body of available genomic and structural data within the kinase family, e.g.~by generating large numbers of starting configurations to be used in highly parallel MD simulation.

We selected all available structures of protein kinase domains (of any species) as templates, for a total of 4433 (\num{398970} target-template pairs).
The templates were derived from 3028 individual PDB entries and encompassed 23 different species, with 3634 template structures from human kinase constructs.
{\color{red}[JDC: Shouldn't we state which options we used and what Uniprot searches we used for templates and targets?  How would someone reproduce what we did here?  Can we communicate this beyond just saying ``here are the scripts''?]}

@@

{\color{red}[JDC: This section looks pretty anemic.  What other kinds of analyses can we do for all the TKs?  There is so much data here!  There must be something neat we can do to examine it, right?]}

@@

To provide a more complete evaluation of the models generated, we have analyzed two example TKs (\emph{Src} and \emph{Abl1}) in detail.
Due to their importance in cancer, these kinases have been the subject of numerous studies, encompassing many different methodologies.
In terms of structural data, a large number of crystal structures have been solved (with or without ligands such as nucleotide substrate or inhibitor drugs), showing the kinases in a number of different conformations.
These two kinases are thus also interesting targets for MSM studies, with one recent study focusing on modeling the states which constitute the activation pathway of Src~\cite{shukla-pande:nature-commun:2014:src-activation-msm}.

Fig.~\ref{figure:superposition} shows a superposition of a set of representative models of \emph{Src} and \emph{Abl1}.
Models were first stratified into three ranges, based on the structure of the sequence identity distribution (Fig.~\ref{figure:sequence-identity-distribution}), then subjected to $k$-medoids clustering to pick three representative models from each sequence identity range.
{\color{red}[JDC: Explain how $k$-medoids clustering was done either here or in figure caption.]}
Each model is colored and given a transparency based on the sequence identity between the target and template sequence.
The figure gives an idea of the variance present in the generated models.
High sequence identity models (in opaque blue) tend to be quite structurally similar, with some variation in loops or changes in domain orientation.

The Abl1 renderings in Fig.~\ref{figure:superposition} indicate one high sequence identity model with a long unstructured region at one of the termini, which was unresolved in the original template structure.
While such models are not necessarily incorrect or undesirable, it is important to be aware of the effects they may have on production simulations performed under periodic boundary conditions, as long unstructured termini can be prone to interact with a protein's periodic image.
Lower sequence identity models (in transparent white or red) indicate much greater variation in all parts of the structure.
We believe the mix of high and low sequence identity models to be particularly useful for methods such as MSM building, which require thorough sampling of the conformational landscape.
The high sequence identity models could be considered to be the most likely to accurately represent true metastable states.
Conversely, the lower sequence identity models could be expected to help push a simulation into regions of conformation space which might take intractably long to reach if starting a single metastable conformation.

To evaluate the models of \emph{Src} and \emph{Abl1} in the context of the published structural biology literature on functionally relevant conformations, we have focused on two residue pair distances thought to be important for the regulation of protein kinase domain activity.
We use the residue numbering schemes for chicken Src (which is commonly used in the literature even in reference to human Src)[CITE: 2SRC, 1Y57] and human Abl1 isoform A[CITE: 2F4J, 2HYY, 2G1T] respectively; the exact numbering schemes are provided in Supporting Information S\ref{si:residue-numbering-schemes}.

Fig.~\ref{figure:src-ref-structures} shows two structures of \emph{Src} believed to represent inactive (PDB code: 2SRC) {\color{red}[CITE: 2SRC]} and active (PDB code: 1Y57) {\color{red}[CITE: 1Y57]} states.
One notable feature which distinguishes the two structures is the transfer of an electrostatic interaction of E310 from R409 (in the inactive state) to K295 (in the active state), brought about by a rotation of the $\alpha$C-helix.
These three residues are also well conserved [CITE Kannan Neuwald JMB 2005], and a number of experimental and simulation studies have suggested that this electrostatic switching process plays a role in a regulatory mechanism shared across the protein kinase family~\cite{shukla-pande:nature-commun:2014:src-activation-msm} [CITE Foda Shan Seeliger Src Nat Commun 2015; Ozkirimli Post Prot Sci 2008].
As such, we have projected the {\bf Ensembler} models for \emph{Src} and \emph{Abl1} onto a space consisting of the distances between these two residue pairs (Fig.~\ref{figure:pair-distances}).
The models show strong coverage of regions in which either of the electrostatic interactions is formed, as well as a wide range of regions inbetween.
We thus expect that such a set of models, if used as starting configurations for highly parallel MD simulation, could greatly aid in sampling of functionally relevant conformational states.
